### 2017-01-16 main.tex
Add interactive essay to abstract
Edits in further work and the rest (Chapter 8)
Addressing feedback, fix demands (Chapter 6)
Edits and tweaks (Chapter 8 and 9 mostly)
Edits pages 0-30

\newcommand{\atimes}{ \textcolor{aclr}{\times} }
\newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq
\newcommand{\aparstr}{ \textcolor{aclr}{\oplus}} % Should be the same as \cseq

@@

\newcommand{\azip}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
\newcommand{\apar}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }

@@

%\include{text/Titlepage} % Main title page

%\cleardoublepage\include{text/Declaration} % Publications from the thesis page

%\cleardoublepage\include{text/Abstract} % Abstract page

@@

\ctparttext{  The computing ecosystem is becoming increasingly heterogeneous and rich. Modern
  programs need to run on a variety of devices that are all different, but provide unique rich
  capabilities. For example, application running on a phone can access the GPS sensor, while
  application running in the cloud can access GPU computing resources. Both diversity and richness
  can only be expected to increase with trends such as the internet of things.

  In this thesis, we argue that the creating programming languages that allow the programmer to
  better work with the environment or \emph{context} in which applications execute is the next
  big challenge for programming language designers.

@@

%\include{text/Chapter01}
%\include{text/Chapter02}
%\include{text/Chapter03}

@@

%\include{text/Chapter04}
%\include{text/Chapter05}
%\include{text/Chapter06}

@@

%\appendix
%\include{text/AppendixA}
%\include{text/AppendixB}

>>>>

\newcommand{\vtimes}{ \mathbin{+\mkern-10mu+} }

\newcommand{\atimes}{ \textcolor{aclr}{\mathbin{+\mkern-10mu+}} }
\newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq
\newcommand{\aparstr}{ \textcolor{aclr}{\oplus}} % Should be the same as \cseq

@@

\newcommand{\azip}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{180}{$\perp$}\mkern-10mu\rotatebox[origin=c]{180}{$\perp$}} } }
\newcommand{\apar}{ \textcolor{aclr}{ \mathbin{\perp\mkern-10mu\perp} } }

@@

\include{text/Titlepage} % Main title page

\cleardoublepage\include{text/Declaration} % Publications from the thesis page

\cleardoublepage\include{text/Abstract} % Abstract page

@@

\ctparttext{  The computing ecosystem is becoming increasingly heterogeneous and rich. Modern
  programs need to run on a variety of devices that are all different, but provide unique rich
  capabilities. For example, an application running on a phone can access the GPS sensor, while
  a service running in the cloud can access GPU computing resources. Both diversity and richness
  can only be expected to increase with trends such as the internet of things.

  In this thesis, we argue that creating programming languages that allow the programmer to
  better work with the environment or \emph{context} in which applications execute is the next
  big challenge for programming language designers.

@@

\include{text/Chapter01}
\include{text/Chapter02}
\include{text/Chapter03}

@@

\include{text/Chapter04}
\include{text/Chapter05}
\include{text/Chapter06}

@@

\appendix
\include{text/AppendixA}
\include{text/AppendixB}

### 2017-01-16 text/Abstract.tex
Edits pages 0-30

We identify a number of program properties that were not connected before, but model some notion
of context. Our examples include tracking different execution platforms (and their versions)
in cross-platform development, resources available in different execution environments (\eg~GPS
sensor on a phone and database on the server), but also more traditional notions such as
variable usage (\eg~in liveness analaysis and linear logics) or past values in
stream-based dataflow programming. Our first contribution is the discovery of the connection
between the above examples and their novel presentation in the form of calculi (\emph{coeffect systems}).
The presented type systems and formal semantics highlight the relationship between different
notions of context.

Our second contribution is the definition of two unified coeffect calculi that capture the common
structure of the examples. In particular, our \emph{flat coeffect calculus} models languages
with contextual properties of the execution environment and our \emph{structural coeffect
calculus} models languages where the contextual properties are attached to the variable usage.
We define the semantics of the calculi in terms of category theoretical structure of an
\emph{indexed comonad} (based on dualisation of the well-known monad structure), use it
to define operational semantics and prove safety result for the calculi.

>>>>

We identify a number of program properties that were not connected before, but model some notion
of context. Our examples include tracking different execution platforms (and their versions)
in cross-platform development, resources available in different execution environments (\eg~GPS
sensor on a phone and database on the server), but also more traditional notions such as
variable usage (\eg~in liveness analysis and linear logics) or past values in
stream-based dataflow programming. Our first contribution is the discovery of the connection
between the above examples and their novel presentation in the form of calculi (\emph{coeffect systems}).
The presented type systems and formal semantics highlight the relationship between different
notions of context.

Our second contribution is the definition of two unified coeffect calculi that capture the common
structure of the examples. In particular, our \emph{flat coeffect calculus} models languages
with contextual properties of the execution environment and our \emph{structural coeffect
calculus} models languages where the contextual properties are attached to the variable usage.
We define the semantics of the calculi in terms of category theoretical structure of an
\emph{indexed comonad} (based on dualisation of the well-known monad structure), use it
to define operational semantics and prove type safety of the calculi.

### 2017-01-16 text/Chapter01.tex
Edits pages 0-30

Before exploring the motivations for to this thesis, we briefly consider two recent practical
concerns that have led to the development of new programming languages. This helps to explain why
context-aware programming is important. The examples are by no means representative, but they
illustrate various kinds of motivations well.

@@

Accessing ``big data'' sources is an example of a more subtle challenge. Initiatives like open
government data\footnote{In the UK, the open government data portal is available at: \url{http://data.gov.uk/}}
certainly make more data available. However, to access the data, one has to parse CSV and Excel files,
issue SQL or SPARQL queries (to query database and the semantic web, respectively).

@@

~\\

@@

This challenge is of the kind that is not easy to see, perhaps because we are so used to
doing things in certain ways that we cannot see their flaws. In this chapter, we aim to
expose such flaws. We look at a number of basic programs that rely on contextual
information, we explain why the currently dominant solutions are inappropriate and then
briefly outline how this thesis solves the problems.

% ==================================================================================================
%
%     ###                  #                    #
%    #   #                 #                    #
%    #       ###   # ##   ####    ###   #   #  ####           ###   #   #   ###   # ##    ###
%    #      #   #  ##  #   #     #   #   # #    #     #####      #  #   #      #  ##  #  #   #
%    #      #   #  #   #   #     #####    #     #             ####  # # #   ####  #      #####
%    #   #  #   #  #   #   #  #  #       # #    #  #         #   #  # # #  #   #  #      #
%     ###    ###   #   #    ##    ###   #   #    ##           ####   # #    ####  #       ###
%                                                                                                                                                                       % ==================================================================================================
% ==================================================================================================

@@

The following sub-sections explore 4 examples in more detail. The examples are chosen to
demonstrate two distinct forms of contexts that are studied in this thesis -- first two are
related to the program environment and the latter two are associated with individual variables of
the program.

@@

Another example related to the previous one is when libraries use meta-programming techniques,
such as LINQ \cite{app-linq,app-linq-essence} or F\# quotations \cite{app-fsharp-metaprog}, to
translate code written in a subset of a host language to some other target language, such as SQL,
OpenCL or JavaScript. For database access, this is a recently developed technique replacing embedded
SQL discussed in the introduction, but it is a more broadly applicable technique for programming
in heterogeneous environments. It lets developers targets multiple runtimes that have limited
execution capabilities.

For example, the following LINQ query written in C\# queries a database and
selects those product names where the first upper case letter is ``C'':
%
\begin{equation*}
\begin{array}{l}
\kvd{var}~\ident{db} = \kvd{new}~\ident{NorthwindDataContext}();\\[0.25em]
\kvd{from}~\ident{p}~\kvd{in}~\ident{db.Products}\\[-0.25em]
\kvd{where}~\ident{p.ProductName.First}(\lambda \ident{c} \rightarrow \ident{Char.IsUpper}(\ident{c})) == \str{C}\\[-0.25em]
\kvd{select}~\ident{p.ProductName};
\end{array}
\end{equation*}
%
This appears as a perfectly valid code and the C\# compiler accepts it. However, when the program
is executed, it fails with the following error:
%
\begin{quote}
\emph{Unhandled Exception: \texttt{System.NotSupportedException}: Sequence operators
not supported for type \texttt{System.String}.}
\end{quote}
%
The problem is that LINQ can only translate a \emph{subset} of normal C\# code. The above snippet
uses the \ident{First} method to iterate over characters of a string, which is not supported. This is
not a technical limitation of LINQ, but a fundamental problem of the approach.

@@

Again, this example demonstrates a more general property. Sometimes, it is desirable to track additional
metadata about variables that are in some ways special. Such metadata can determine how the variables
can be used. Here, \ident{name} comes from the user input. This information about the value should
be propagated to \ident{query}. The \ident{SqlCommand} object should then require arguments that can
not directly contain user input (in an unchecked form).

@@

Consider a simple programming language with arrays where $n^{\mathit{th}}$ element of an
array \ident{arr} is accessed using $\ident{arr}[n]$. We focus on writing stencil computations
(such as image blurring, Conway's game of life or convolution) where all arrays are of
the same size and the system provides a \emph{cursor} pointing to a current location in the
stencil. We assume that the keyword \kvd{cursor} returns the current location in the stencil.

@@

As an example, consider a news reader app consisting of a server-side component (which stores the
news in an SQL database) and a number of clients applications for popular platforms (Android, Windows
Phone, etc.). A simplified code excerpt that might appear somewhere in the implementation is shown in
Figure~\ref{fig:introduction-sample}.

We assume that the language supports cross-compilation and splits the single program into three
components: one for the server-side and two for the client-side, for iPhone and Windows platforms,
respectively. The cross-compilation could be done in a way similar to Links \cite{app-distributed-links},
but we do not require explicit annotations specifying the target platform.

@@

Furthermore, integrating contextual information into the type system can provide information for
modern developer tools. For example, many editors for F\# display inferred types when placing mouse
pointer over an identifier. For \ident{fetchLocalNews}, the tip could appear as follows:
%
\definecolor{tooltipy}{rgb}{0.98,0.97,0.76}
\definecolor{tooltipb}{rgb}{0.7,0.7,0.7}
\vspace{0.5em}
\begin{equation*}
\hspace{5em}
\begin{array}{l}
\ident{fetchLocalNews}\\
\hspace{1em} \fcolorbox{tooltipb}{tooltipy}{
	\introfuncblack{unit}{gps, rpc}{(news~list)~async}
}
\end{array}
\end{equation*}
\vspace{0.5em}

@@

The concrete syntax used here is just for illustration. Furthermore,
some information could even be mapped to other visual representations -- for example, differently
coloured backgrounds for platform-specific functions. The key point is that the type provides a
number of useful information:

@@

\item The \ident{gpsLocation} function accesses the GPS sensor and since we call it in
  from \texttt{fetchLocalNews}, this function also requires GPS (the requirement is propagated
  automatically).

\item We can compile the program for two client-side platforms - the entry points are \ident{iPhoneMain}
  and \ident{windowsMain} and require iOS and Windows user-interface libraries, together with
  GPS and the ability to perform remote calls over the network.
\end{itemize}

@@

\paragraph{Type systems.}
A type system is a form of static analysis that is usually specified by \emph{typing judgements}
such as $\Gamma \vdash e : \tau$. The judgement specifies that, given some variables described by
the context $\Gamma$, the expression $e$ has a type $\tau$.
The variable context $\Gamma$ is necessary to determine the type of expressions. Consider an
expression $x + y$. In many languages, including Java, C\# and F\#, the type could be \ident{int},
\ident{float} or \ident{string}, depending on the types of the variables. For example,
the following is a valid typing judgement in F\# \cite{app-fsharp}:
%
\begin{equation*}
x\!:\!\ident{int},~y\!:\!\ident{int} \vdash x+y : \ident{int}
\end{equation*}
%
This judgement assumes that the type of both $x$ and $y$ is \ident{int} and so the result must also
be \ident{int}. In F\#, the expression would also be typeable in a context $x\!:\!\ident{string},~y\!:\!\ident{string}$,
but not, for example, in a context where $x$ has a type \ident{int} and $y$ has a type \ident{string}.

@@

Effect systems are an established idea, but they are suitable only for tracking properties of a
certain kind. They can be used for properties that describe how programs \emph{affect} the environment.
For context-aware languages, we instead need to track what programs \emph{require} from the environment.

@@

The systems for tracking of context demands developed in this thesis are inspired by the idea
of effect systems. To demonstrate our approach, consider the following call from the sample program
shown earlier -- first using standard ML-like type system:
%
\begin{equation*}
\ident{password}\!:\!\ident{string},~\ident{cmd}\!:\!\ident{string}
  \vdash \ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{equation*}
%
The expression queries a database and gets back a list of news values as the result. Recall from the
earlier discussion that there are two contextual information that are desirable to track for this
expression. First, the call to the \ident{query} primitive requires \emph{database access}. Second,
the \ident{password} argument needs to be marked as \emph{sensitive value} to avoid sending it over
an unsecure network connection. The \emph{coeffect systems} developed in this thesis capture this
information in the following way (we slightly refine the notation later):
%
\begin{equation*}
\begin{array}{l}
 \coctx{( \ident{password}\!:\!\coctx{\ident{string}}{\hspace{-0.75em}\ident{\cclrd{\footnotesize sensitive}}},
   \ident{cmd}\!:\!\ident{string})}{ \aclrd{\{ \ident{database} \}}}  \\[-0.1em]
 \hspace{3em} \vdash\ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{array}
\end{equation*}
%
Rather than attaching the annotation to the \emph{resulting type}, we attach them to the variable
context $\Gamma$. In other words, coeffect systems do not keep track just of the variables available
in the context -- they also capture detailed information about the execution environment.
In the above example, the system tracks metadata about the variables and annotates \ident{password}
as sensitive. Furthermore, it tracks requirements about the execution environment, for example,
that the execution requires an access to database.

@@

It is well-known fact that \emph{effects} correspond to \emph{monads} and languages such as Haskell
use monads to provide a limited form of effect system. An interesting observation made in this thesis
is that \emph{coeffects}, or systems for tracking contextual information, correspond to the category
theoretical dual of monads called \emph{comonads}. The details are explained throughout the thesis.

@@

The key claim of this thesis is that programming languages need to provide better ways of capturing
how programs rely on the context, or execution environment, in which they execute. This chapter
shows why this is an important problem. We looked at a number of properties related to context that
are currently handled in ad-hoc and error-prone ways. Next, we considered the properties in a
simplified, but realistic example of a client/server application for displaying local news.

@@

\item In Chapter~\ref{ch:impl}, we use the translation as a basis for a prototype
  implementation of three simple context-aware programming languages. Our implementation serves as
  an empirical evidence showing how coeffects simplify programming with context and provide
  additional safety guarantees. We present the implementation in the format of an \emph{interactive
  essay} (\url{http://tomasp.net/coeffects}) that encourages active reading and exploration of
  various aspects of the theory.

@@

\noindent
If there is a one thing that the reader should remember from this thesis, it is the fact that
there is a unified notion of \emph{context}, capturing many common scenarios in programming,
and that programming language designers need to provide ways for working with this context (using
\emph{coeffects} or not). This greatly reduces the number of distinct concepts that software
developers need to keep in mind of when building applications for the rich and diverse
execution environments of the future.

>>>>

Before exploring the motivations for to this thesis, we briefly consider two recent practical
concerns that have led to the development of new programming languages. This helps to explain why
context-aware programming is important. The examples are by no means exhaustive, but they
are representative of the various kinds of motivations.

@@

Accessing ``big data'' is an example of a more subtle challenge. Initiatives like open
government data make more data available, but to access the data, one has to parse CSV and 
Excel files, issue SQL or SPARQL queries (to query database and the semantic web, respectively).

@@

@@

This challenge is of the kind that is not easy to see, perhaps because we are so used to
doing things in certain ways that we cannot see their flaws. In this chapter, we aim to
expose such flaws. We look at a number of basic programs that rely on contextual
information, we explain why the currently dominant solutions are inappropriate.

\paragraph{The thesis.}
The key claim of this thesis is that the theory of \emph{coeffects} that we develop provides
a unified way of capturing how programs rely on the context, or execution environment, 
in which they execute. Perhaps most importantly, coeffects provide a single conceptual framework 
that can capture a wide range of previously disconnected notions of context-awareness.

Thanks to coeffects, programming languages for rich and diverse execution environments will
be able to offer a simple, safe and unified programming model for interacting with the 
environment. In this thesis, we develop the theoretical foundations for such languages and
we develop a prototype implementation that illustrates the benefits of coeffects, but also 
serves as a rich guide for future langauge implementers.

% ==================================================================================================
%
%     ###                  #                    #
%    #   #                 #                    #
%    #       ###   # ##   ####    ###   #   #  ####           ###   #   #   ###   # ##    ###
%    #      #   #  ##  #   #     #   #   # #    #     #####      #  #   #      #  ##  #  #   #
%    #      #   #  #   #   #     #####    #     #             ####  # # #   ####  #      #####
%    #   #  #   #  #   #   #  #  #       # #    #  #         #   #  # # #  #   #  #      #
%     ###    ###   #   #    ##    ###   #   #    ##           ####   # #    ####  #       ###
%                                                                                                                                                                       
% ==================================================================================================

@@

The following sub-sections explore 4 examples in more detail. The examples are chosen to
demonstrate two distinct forms of contexts that are studied in this thesis -- the first two are
related to the program environment and the latter two are associated with individual variables of
the program.

@@

Another example related to the previous one is when libraries use meta-programming techniques,
such as LINQ \cite{app-linq,app-linq-essence} or F\# quotations \cite{app-fsharp-metaprog}, to
translate code written in a subset of a host language to some other target language, such as SQL,
OpenCL or JavaScript. For database access, this is a recently developed technique replacing embedded
SQL discussed in the introduction, but it is a more broadly applicable technique for programming
in heterogeneous environments. It lets developers target multiple runtimes that have limited
execution capabilities.

For example, the following LINQ query written in C\# queries a database and
selects those product names where the first upper case letter is ``C'':
%
\begin{equation*}
\begin{array}{l}
\kvd{var}~\ident{db} = \kvd{new}~\ident{NorthwindDataContext}();\\[0.25em]
\kvd{from}~\ident{p}~\kvd{in}~\ident{db.Products}\\[-0.25em]
\kvd{where}~\ident{p.ProductName.First}(\lambda \ident{c} \rightarrow \ident{Char.IsUpper}(\ident{c})) == \str{C}\\[-0.25em]
\kvd{select}~\ident{p.ProductName};
\end{array}
\end{equation*}
%
This appears as a perfectly valid code and the C\# compiler accepts it. However, when the program
is executed, it fails with the following error:
%
\begin{quote}
\emph{Unhandled Exception: \texttt{System.NotSupportedException}: Sequence operators
not supported for type \texttt{System.String}.}
\end{quote}
%
The problem is that LINQ can only translate a \emph{subset} of normal C\# code. The above snippet
uses the \ident{First} method to iterate over characters of a string, which is not supported. This is
not a technical limitation of LINQ, but a fundamental problem of an approach where code written 
using libraries for one execution environment is translated and run in another, more limited, 
execution environment.

@@

Again, this example demonstrates a more general property. Sometimes, it is desirable to track additional
metadata about values that are in some ways special. Such metadata can determine how the values
can be used. Here, the value stored in \ident{name} comes from the user input. This information about 
the value should be propagated to \ident{query}. The \ident{SqlCommand} object should then require 
arguments that can not directly contain user input.

@@

Consider a simple programming language with arrays where the $n^{\textit{th}}$ element of an
array \ident{arr} is accessed using $\ident{arr}[n]$. We focus on writing stencil computations
(such as image blurring, Conway's game of life or convolution) where all arrays are of
the same size and the system provides a \emph{cursor} pointing to a current location in the
stencil. We assume that the keyword \kvd{cursor} returns the current location in the stencil.

@@

As an example, consider a news reader app consisting of a server-side component (which stores the
news in an SQL database) and a number of clients applications for popular platforms (iPhone, Windows
Phone, etc.). A simplified code excerpt that might appear somewhere in the implementation is shown in
Figure~\ref{fig:introduction-sample}.

We assume that the language supports cross-compilation and splits the single program into three
components: one for the server-side and two for the client-side, for iPhone and Windows Phone,
respectively. The cross-compilation could be done in a way similar to Links \cite{app-distributed-links},
but we do not require explicit annotations specifying the target platform.

@@

Furthermore, integrating contextual information into the type system can provide information for
modern developer tools. For example, many editors for F\# display inferred types when placing the 
mouse pointer over an identifier. For \ident{fetchLocalNews}, the tip could appear as follows:
%
\definecolor{tooltipy}{rgb}{0.98,0.97,0.76}
\definecolor{tooltipb}{rgb}{0.7,0.7,0.7}
\vspace{0.5em}
\begin{equation*}
\hspace{5em}
\begin{array}{l}
\ident{fetchLocalNews}\\
\hspace{1em} \fcolorbox{tooltipb}{tooltipy}{
	\introfuncblack{unit}{gps, rpc}{(news~list)~async}
}
\end{array}
\end{equation*}
\vspace{0.5em}

@@

The concrete syntax is just for illustration and
some information could even be mapped to other visual representations -- for example, differently
coloured backgrounds for platform-specific functions. The key point is that the type provides a
number of pieces of useful information:

@@

\item The \ident{gpsLocation} function accesses the GPS sensor and since we call it 
  from \texttt{fetchLocalNews}, this function also requires GPS (the requirement is propagated
  automatically).

\item We can compile the program for two client-side platforms -- the entry points are \ident{iPhoneMain}
  and \ident{windowsMain} and require iOS and Windows user-interface libraries, together with
  GPS and the ability to perform remote calls over the network.
\end{itemize}

@@

\paragraph{Type systems.}
A type system is a form of static analysis that is usually specified by \emph{typing judgements}
such as $\Gamma \vdash e : \tau$. The judgement specifies that, given some variables described by
the context $\Gamma$, the expression $e$ has a type $\tau$.
The variable context $\Gamma$ is necessary to determine the type of expressions. Consider an
expression $x + y$. In many languages, including Java, C\# and F\#, the type could be \ident{int},
\ident{float} or \ident{string}, depending on the types of the variables. For example,
the following is a valid typing judgement in F\# \cite{app-fsharp}:
%
\begin{equation*}
x\!:\!\ident{int},~y\!:\!\ident{int} \vdash x+y : \ident{int}
\end{equation*}
%
This judgement assumes that the type of $x$ and $y$ is \ident{int} and so the result must also
be \ident{int}. The expression might also be typeable in a context $x\!:\!\ident{string},~y\!:\!\ident{string}$,
but not in a context where types of $x$ and $y$ do not match.

@@

Effect systems are an established idea, but they are suitable only for tracking properties of a
certain kind. They can be used for properties that describe how programs \emph{affect} the environment.
For context-aware languages, we instead need to track what programs \emph{require} from the 
environment. This intuitive distinction is made more precise in Section~\ref{sec:semantics-related}.

@@

The systems for tracking of context demands developed in this thesis are inspired by the idea
of effect systems. To demonstrate our approach, consider the following call from the sample program
shown earlier -- first using standard ML-like type system:
%
\begin{equation*}
\ident{password}\!:\!\ident{string},~\ident{cmd}\!:\!\ident{string}
  \vdash \ident{query}(\ident{cmd}, \ident{password}, \kvd{time}) : \ident{news list}
\end{equation*}
%
The expression queries a database and gets back a list of news based on the spefieid \kvd{time}. 
There are three pieces of contextual information that are desirable to track for this expression. 
First, the call to the \ident{query} primitive requires \emph{database access}. Second, 
the \ident{password} argument needs to be marked as \emph{sensitive value} to avoid sending it over
an unsecure network connection. Third, the expression uses a special construct \kvd{time}, which
requires access to \emph{system clock}. The \emph{coeffect systems} developed in this thesis 
capture this information in the following way (we slightly refine the notation later):
%
\begin{equation*}
\begin{array}{l}
 \coctx{( \ident{password}\!:\!\coctx{\ident{string}}{\hspace{-0.75em}\ident{\cclrd{\footnotesize sensitive}}},
   \ident{cmd}\!:\!\ident{string})}{ \aclrd{\{ \ident{database}, \ident{clock} \}}}  \\[-0.1em]
 \hspace{3em} \vdash\ident{query}(\ident{cmd}, \ident{password}, \kvd{time}) : \ident{news list}
\end{array}
\end{equation*}
%
The typing judgement includes an additional annotation that now captures 
contextual requirements of the expression. The annotation is attched to the variable context 
$\Gamma$. As discussed later (Section~\ref{sec:flat-calculus-understanding}), this better reflects
how contextual information are used in the type system, but it also matches the semantics 
(Section~\ref{sec:semantics-theory}). Rather than attaching additional information to the
\emph{resulting type}, we attach them to the variable context $\Gamma$.

In other words, the context in coeffect systems consists of the available variables, but also
tracks detailed information about the execution environment.
In the above example, the system tracks metadata about the variables and annotates \ident{password}
as sensitive and it tracks requirements about the execution environment, for example,
that the execution requires clock.

@@

\paragraph{Context demands and lambda abstraction.}
The difference between effects and coeffects becomes apparent when we consider lambda abstraction.
Given an effectful expression such as $\ident{print}~\str{hi}$, the function $\lambda x.\ident{print}~\str{hi}$
is an effect free value that delays all effects. The message is printed when we run the function,
but not when the function is declared.

Contextual properties do not follow this pattern. As discussed in 
Section~\ref{sec:applications-structure-lam}, context demands cannot always be delayed. For 
example, consider a context-aware expression $\kvd{time}$ that requires access to system clock.
If we create this function on a server in a distributed system, send it to a client and then
run it, the function body can access the clock in two ways. It can use the clock of the server (when the
function was created), or it can use the clock of the client (when the function was executed).
We return to this example in Section~\ref{sec:applications-flat-distr}.

Another important difference between effects and coeffects becomes apparent when we consider their
semantics. It is well-known fact that \emph{effects} correspond to \emph{monads} and languages such as Haskell
use monads to provide a limited form of effect system. An interesting observation made in this thesis
is that \emph{coeffects}, or systems for tracking contextual information, correspond to the category
theoretical dual of monads called \emph{comonads}. The details are explained throughout the thesis.

@@

This chapter shows why capturing how programs rely on the context, or execution environment, in 
which they execute is an important problem. We looked at a number of properties related to context 
that are currently handled in ad-hoc and error-prone ways. Next, we considered the properties in a
simplified, but realistic example of a client/server application for displaying local news.

@@

\item In Chapter~\ref{ch:impl}, we use the translation as a basis for a prototype
  implementation of three simple context-aware programming languages. We use a format of web-based
  \emph{interactive essay} (available at \url{http://tomasp.net/coeffects}), which shows how 
  coeffects simplify programming with context by allowing the reader to write and run simple 
  context-aware programs, but also explore the theory behind the implementation including the 
  typing derivation and the translation.

@@

\noindent
If there is a one thing that the reader should remember from this thesis, it is the fact that
there is a unified notion of \emph{context}, capturing many common scenarios in programming,
and that programming language designers need to provide ways for working with this context.
This greatly reduces the number of distinct concepts that software
developers need to keep in mind of when building applications for the rich and diverse
execution environments of the future.
